---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
```{r Load libraries and remove scientific notation, echo=FALSE,results='hide'}
    require("ggplot2")
    options(scipen = 999)
```

## Loading and preprocessing the data
The dataset for the current project can be found at [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip).

Extract `activity.csv` from `repdata_data_activity.zip` which is downloaded from [Activity Monitoring Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) if it is not already present in the working directory.
```{r Download & Extract Data}
    url            <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    compressedName <- "repdata_data_activity.zip"
    datasetName    <- "activity.csv"
    
    ##  Check for the presence of compressedName & datasetName. If not present, download.
    if (!(file.exists(compressedName) | file.exists(datasetName))) 
        download.file(url, compressedName)
    ##  Check for presence of datasetName. If not present, extract.
    if (!(file.exists(datasetName))) unzip(compressedName)
    
```

Read the data from `activity.csv` into `activity` (dataFrame).
```{r Load Data into a Data Frame}
    activity        <- read.csv(datasetName)
    activity$date   <- as.Date(activity$date) 
```

## What is mean total number of steps taken per day?
Aggregate the given data into a data frame with total steps and date as columns named  `stepsPerDay`. Ignore any data with `NA`.
```{r Find the Total Steps per day}
    stepsPerDay <- aggregate(steps ~ date, activity, sum)
```

Calculate the `mean` and the `median` of the total number of steps.
```{r Mean and Median of Total number of steps taken each day}
    meanDailySteps      <- mean(stepsPerDay$steps)
    medianDailySteps    <- median(stepsPerDay$steps) 
```

Plot a `histogram` for `Total steps taken per day`.
```{r Histogram of Total steps taken each day}
    graph <- ggplot(stepsPerDay, aes(x = steps)) + geom_histogram(bins = 5)
    graph <- graph + geom_vline(xintercept = meanDailySteps, size = 2)
    graph <- graph + geom_vline(xintercept = medianDailySteps, size = 2)
    print(graph)
```

The mean total number of steps taken per day is `r meanDailySteps`.
The median of the same is `r medianDailySteps`.

## What is the average daily activity pattern?
Aggregate the given data into a data frame with average steps and interval as columns named  `avgStepsPerInt`. Ignore any data with `NA`.
```{r Find the Average Steps per Interval}
    avgStepsPerInt <- aggregate(steps ~ interval, activity, mean)
```

Calculate the `max` of the average steps.
```{r Maximum of average steps taken each day}
    rowindex    <- which.max(avgStepsPerInt$steps)
    maxAvgSteps <- avgStepsPerInt[rowindex,]$steps
    maxInterval <- avgStepsPerInt[rowindex,]$interval
```

Plot a `line graph` for `Average steps taken per Interval`.
```{r line graph of average steps taken per interval}
    graph <- ggplot(avgStepsPerInt, aes(x = interval, y = steps)) + geom_line()
    graph <- graph + geom_vline(aes(xintercept = maxInterval), colour="#BB0000")
    print(graph)
```

The maximum average steps taken in an interval is `r maxAvgSteps`.
The maximum average steps taken in the interval `r maxInterval`.

## Imputing missing values
Calculate the number of rows containing `NA`.
```{r Find the Total Missing Steps}
    nMissingRows <- sum(is.na(activity$steps))
```

There are `r nMissingRows` in the given data `activity`.

Create a new `dataFrame` to store the new imputed Data `imputedActivity`
```{r create dataFrame imputedActivity}
    imputedActivity <- activity
```
Fill in the `NA` spaces with the average steps taken in that interval.
```{r Fill in the missing data from the data set activity}
    naIntervals <- activity[is.na(activity$steps),3]
    imputedActivity[is.na(imputedActivity$steps),1] <- avgStepsPerInt[naIntervals%/%5+1,2]
```

Aggregate the given data into a data frame with total steps and date as columns named  `stepsPerDay`. Ignore any data with `NA`.
```{r Find the Total Steps per day using imputed data}
    imputedStepsPD <- aggregate(steps ~ date, imputedActivity, sum)
```

Calculate the `mean` and the `median` of the total number of steps.
```{r Imputed Mean and Median of Total number of steps taken each day}
    imputedMeanDS      <- mean(imputedStepsPD$steps)
    imputedMedianDS    <- median(imputedStepsPD$steps) 
```

Plot a `histogram` for `Total steps taken per day`.
```{r Histogram of Total imputed steps taken each day}
    graph <- ggplot(imputedStepsPD, aes(x = steps)) + geom_histogram(bins = 5)
    graph <- graph + geom_vline(xintercept = imputedMeanDS, size = 2)
    graph <- graph + geom_vline(xintercept = imputedMedianDS, size = 2)
    print(graph)
```

The imputed mean total number of steps taken per day is `r imputedMeanDS`.
The imputed median of the same is `r imputedMedianDS`.

Both the mean and median calculated from the imputed values are lower by a small amount.

The mean is lower by `r meanDailySteps - imputedMeanDS`.
The median is lower by `r medianDailySteps - imputedMedianDS`.

## Are there differences in activity patterns between weekdays and weekends?
Divide the days into weekdays and weekends and add the column dayofweek to `activity`.
```{r Add factor weekends and weekdays}
    logicDayofWeek <- weekdays(imputedActivity$date) %in% c("Sunday", "Saturday")
    imputedActivity$dayofweek <- "weekday"
    imputedActivity[logicDayofWeek,4] <- "weekend"
    imputedActivity$dayofweek <- as.factor(imputedActivity$dayofweek)
```

Aggregate the given data into a data frame with average steps and interval as columns named  `avgStepsPerInt`. Ignore any data with `NA`.
```{r Find the Average Steps per Interval and weekday}
    avgStepsPerIntW <- aggregate(steps ~ interval + dayofweek, imputedActivity, mean)
```

Plot a `line graph` for `Average steps taken per Interval during weekdays and weekends`.
```{r line graph for avg. steps taken per int. during weekdays and weekends}
    graph <- ggplot(avgStepsPerIntW, aes(x = interval, y = steps)) + geom_line()
    graph <- graph + facet_grid(dayofweek~.)
    print(graph)
```

#### End of the Document